
---
title: "TimeSeries Forecasting"
author: "Siva Chandrasekar"
date: "November 19, 2017"
output: html_document
---

```{r Loading Required Libraries, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

options(scipen = 999)
library(dplyr)
library(fpp)
library(plotly)
library(dygraphs)

```


# TIME SERIES FORECASTING
This repository for Forecasting Tutorial using Rob Hyndmann's source

This project has all the codes that will be described in Rob Hyndmann's tutorial available at https://www.otexts.org/fpp/1/1

QUESTIONS TO BE ASKED BEFORE FORECASTING:

- Are there consistent patterns?
- Is there a significant trend?
- Is seasonality important?
- Are there any outliers in the data that needs expertise's knowledge?

CHOOSING AND FITTING MODELS:

- Always run different models to predict the forecast, and choose a best model out of the predicted models. Severl tests are available to measure the error's, minimize the errors that will be discussed eventually.

```{r Reading Sample Time Series dataset, results='hide'}

data <- read.csv("~/R Projects/Forecasting-Tutorial/dataSet/SampleData.csv", header = T)
tsdata <- data$tsValue
tsdata <- ts(tsdata, start = c(2010,9), end = c(2017,8), frequency = 12)
head(tsdata)


```

## GRAPHICS:

```{r A simple sample plot}
load("~/R Projects/Forecasting-Tutorial/dataSet/melsyd.rda")

plot(melsyd[,"Economy.Class"], 
     main="Economy class passengers: Melbourne-Sydney",
     xlab="Year",ylab="Thousands")



```

Looking at the plot for the Economy class, the following are evident at a first glance:

- No passengers have travelled during part of the period in 1989 - This is due to some industrial dispute (SME input)

- A period of reduced load in 1992

- A large increase in passenge load in end of 1991

- There are some large dips at the start of each year (holiday season)

- Long term flucutations: increasing in 1987, then decreasing in 1989, then again increasing through 1991, 1992

All the above observations should be accounted when building a model

```{r Another Sample Plot}

dygraph(tsdata, main = "Total Sales", ylab = "Sales in Mn") %>% 
  dyRangeSelector() %>%
  dyAxis("x", label = "Period") %>%
  dyAxis("y", axisLabelFontSize = 10) %>%
  dySeries(label = "Total Sales in USD") %>%
  dyOptions(drawPoints=T, pointSize=3, maxNumberWidth = 20)

```


A simpler time series data is available below:

```{r Plot of a Simpler Time Series data} 

load("~/R Projects/Forecasting-Tutorial/dataSet/a10.rda")

plot(a10,
ylab="$ million",
xlab="Year", 
main="Antidiabetic drug sales")

```

The above graph of Antibiotic drug sales shows the following characteristics:

- Clear increasing trend

- seasonal pattern exists that increases as the level of the series increases

- A dip at the end of each year (due to govt subsidization as provided by SME)

Any model used to predict the antibiotic sales should take in to account on the seasonality, and the slow but steady raising trend.

## SEASONAL PLOT AND MONTH PLOT

These plots are mainly used to visualize the data by season or a cycle which provides better idea to the seasonality behind the data

```{R Seasonplot and monthplot using sample dataset}

load("~/R Projects/Forecasting-Tutorial/dataSet/a10.rda")

seasonplot(a10,
ylab="$ million",
xlab="Year", 
main="Antidiabetic drug sales")

seasonplot(tsdata,
     main = "Sample TS Data",
     xlab="Year",
     ylab="Sample Value",
     year.labels=TRUE, 
     year.labels.left=TRUE,
     col = 1:8,
     pch=8)

monthplot(a10,ylab="$ million",xlab="Month",xaxt="n",
          main="Seasonal deviation plot: antidiabetic drug sales")
axis(1,at=1:12,labels=month.abb,cex=0.8)

monthplot(tsdata,
          main = "Sample TS Data",
     xlab="Year",
     ylab="Sample Value")

```

## SCATTER PLOT

Scatter plots are useful to find the **relations betweek variables**. These plots can also be useful to find the strength of the relationship between a predictor variable and predicted variable.

```{r Scatter Plot Sample}

plot(fuel[,5], fuel[,8], xlab="City mpg", ylab="Carbon footprint")

```

In a scatterplot, *jitter function* is useful to avoid any overlapping of the data points.

```{r Scatter Plot using Jitter function}

plot(jitter(fuel[,5]), jitter(fuel[,8]), xlab="City mpg", ylab="Carbon footprint")

```

**Scatter Plot Matrices** are useful when there are more than two variables for which the relationship is to be determined. 

```{r Pairs plot for comparing multiple continuous variable relationship}

pairs(fuel[,-c(1:2,4,7)], pch=10)

```


Note that scatter plots should be used with *continuous variables* for finding the relationship. Ordinal or factors do not provide much information.

# NUMERICAL SUMMARIES

**Percentiles** are very useful for describing the distribution of the data. 

```{r Finding distribution of data using percentiles}

quantile(fuel$Carbon, probs = .1)
fuel$Carbon
IQR(fuel$Carbon)

quantile(fuel$Carbon, probs = .75) - quantile(fuel$Carbon, probs = .25)

```


**Standard Deviation** is also used to find the spread or the distribution of the data.

```{r Standard Deviation}

sd(fuel$Carbon)

```

## BIVARIATE STATISTICS

Most common bivariate statistics is *correlation coefficient*. This is useful to measure the relationship between two variables. The expression can be explained as below:

 $$r =\frac{\sum_ ( X_i - \bar{X} ) (Y_i - \bar{Y})}{\sqrt{\sum_ ( X_i - \bar{X} )^2} \sqrt{\sum_ ( Y_i - \bar{Y} )^2}}$$

The value of correlation coefficient lies between -1 and +1, the positive values indicating linear relationship, whereas the negative values represent non-linear relationship.

## AUTO CORRELATION FACTOR

**Auto Correlation Factor (ACF)** are used in Time Series for finding the relation between lagged values of time series. For instance, $r_1$  measures the relation between time $y_t$ and $y_{t_-1}$, and $r_2$ measures the relationship between $y_t$ and $y_{t-2}$

```{r Auto Correlation Factor}

lag.plot(tsdata, lags = 12, pch=19, do.lines = F, col = 3)

```

*ACF* can be written as:

$$r_k = \frac{\sum_{t=K+1}^T (y_t - \bar{y}) (y_{t-k} - \bar{y})}{\sum_{t=1}^T (y_t - \bar{y})^2} $$

The ACF between different lags are normally plotted to form the autocorrelation graphs or the correlogram. A sample of ACF is given below:

```{r ACF Plots}

beer2 <- window(ausbeer, start=1992, end=2006-.1)

plot(beer2)

acf(beer2)

acf(tsdata)

```
     
In the beer plot, we can see that there are spikes for every 4 lags, and troughs for every two lags before the spikes. This can also mean that the data has some kind of seasonality every four months

## WHITE NOISE

A scenario where a time series data shows no correlation. 

```{r White Noise Plots}
set.seed(30)
x <- ts(rnorm(50))
plot(x, main="White noise")
acf(x)
```


In white noise, the expectation is that the ACF is close to zero. In other words, 95% of the ACF lies within $\pm \frac{2}{\sqrt{T}}$ where T is the length of time series

# SIMPLE FORECASTING METHODS